Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-01 demystify.tex
updates

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:non-us-differences}), which might require MS (\autoref{sec:phd-vs-ms}) first.

@@

\begin{table}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\paragraph{LoR from Famous People} Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This applies to administrative (e.g., chairs, directors, and deans) also (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

\subsection{Waiving Your Right}\label{sec:waive-right}

When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should \emph{always} waive your right.

@@

\subsection{Help your LOR Writer}\label{sec:help-your-LOR-writers}

As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\subsection{Remind Your Writers}\label{sec:remind-writers}

After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.  Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

\subsection{Thank Your Writers}\label{sec:thank-writers}

After your application is submitted and your writers have submitted their letters (i.e., the wait begins), you should send a quick thank you note, which serves both as an acknowledgement that you know they have submitted the letters and as an appreciate for their help.  You should also let them know the outcome of your application, regardless of whether you are admitted or not.  In addition to being a common courtesy, this can also help maintain a good relationship with your writers, which can be useful in the future (e.g., if you need another letter for another round of applications or for a job reference).

@@

If a student asks me to write letter for them, I will agree as I believe it is part of my job. However, I will ask them to waive their right to see the letter (\autoref{sec:waive-right}), and will not write if they do not do so.

@@

For students who I know well, i.e., strong letters, then writing will take a lot longer as it will be personalized.  While I have a general template of such strong letters, it still can take me an hour or more to write such a letter.  I often ask the students to provide information (\autoref{sec:help-your-LOR-writers}) and what they think I should highlight in the letter. They can also provide me their SOP (if they already have written one) so I can complement what they say with my own perspective.

>>>>

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer compared to other countries (\autoref{sec:non-us-differences}), which might require MS (\autoref{sec:phd-vs-ms}) first.

@@

\begin{table}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\center
\small
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsection{LoR from Famous People}

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This also applies to administrative, e.g., chairs, directors, and deans (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

\subsection{What can you do?}\label{sec:lor-what-to-do}

While you cannot control what your letter writers write, there are multiple things you can do to help them write a strong letter for you.

\paragraph{Waiving Your Right} When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should \emph{always} waive your right.

@@

\paragraph{Help your LOR Writer} As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\paragraph{Remind Your Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers are not even given any deadline.  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

\paragraph{Thank Your Writers} After your application is submitted and your writers have submitted their letters (i.e., the wait begins), you should send a quick thank you note, which serves both as an acknowledgement that you know they have submitted the letters and as an appreciate for their help.  You should also let them know the outcome of your application, regardless of whether you are admitted or not.  In addition to being a common courtesy, this can also help maintain a good relationship with your writers, which can be useful in the future (e.g., if you need another letter for another round of applications or for a job reference).

@@

If a student asks me to write letter for them, I will agree as I believe it is part of my job. However, I will ask them to waive their right to see the letter (\autoref{sec:lor-what-to-do}), and will not write if they do not do so.

@@

For students who I know well, i.e., strong letters, then writing will take a lot longer as it will be personalized.  While I have a general template of such strong letters, it still can take me an hour or more to write such a letter.  I often ask the students to provide information (\autoref{sec:lor-what-to-do}) and what they think I should highlight in the letter. They can also provide me their SOP (if they already have written one) so I can complement what they say with my own perspective.
